Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2023-12-13 article.qmd
Spellcheck

Chemical Reaction Networks (CRNs)
are a fundamental concept of modelling in numerous fields including systems biology, biochemistry, chemical engineering and epidemeology.
They are comprised of a set of chemical species or biological entities and represntation of complex interaction and transformation between them through a series reactions.
These systems can be modelled and simulated through multiple approaches:
deterministic Ordinary Differential Equations (ODEs) to model macroscopic behavior,
Stochastic Differential Equations (SDEs) to model microscopic fluctuations,
and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
Instead of directly writing the equations for each of these formulations
which is error-prone and difficult to reuse,
these models can be written in a higher-level description
that can be compiled for these different types of simulations.

Several tools already exist for defining and simulating CRNs.
BioSimulators.org [@shaikhBioSimulatorsCentralRegistry2022],
a registry of simulation tools,
list at least 15 Python software including COPASI [@hoopsCOPASICOmplexPAthway2006], Tellurium [@choiTelluriumExtensiblePythonbased2018] and pySB [@lopezProgrammingBiologicalModels2013]. 
COPASI is a standalone software with a GUI for defining and simulating models.
It's widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow advanced scripting.
Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to simulate models.
PySB is a Python library
that created a DSL using standard Python to define models
which are then compiled to ODE using a Perl library called BioNetGen [@harrisBioNetGenAdvancesRulebased2016].

@@

Another challenge is the way models are written.
Many tools use a Domain Specific Language (DSL) for coding,
and support the System Biology Markup Language (SBML) as an exchange format [@huckaSBMLL3V2],
as direct SBML coding is impractical.
A DSL allows reuse of code in different programming environments,
but it will not be recognized by default in Integrated Development Environments (IDEs).
Therefore DSLs cannot provide the development experience such as syntax highlighting, code completion, refactoring, and static analysis,
unless such support is specifcally developed.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
An extension for Visual Studio Code was developed, but its maintenance could be demanding task for the systems biology community.
PySB, using Python’s dynamic nature, created a DSL within Python.
By default, it uses global state to create species and parameters,
without assigning them to Python variables
or adding them explicitly to the model,
but this approach is not fully compatible with IDEs,
affecting the development experience.

To overcome these limitations,
we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows defining differential equation systems,
while SimBio builds on it for defining reaction networks.
They are focused on providing an ergonomic experience to end-users by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
Since they are coded in Python,
every part from model definition to simulation can be extended from Python syntax.
Being ?first-ever? pure Python packages for systems modelling,
they offer extensive extensibility,
from simple tasks like reusing integrators defined in other packages,
to complex ones like altering the compilation process to leverage some structure in the equations (For example...).
The models built using these packages can be introspected to create other representations,
such as graphs connecting species and/or reactions,
or tables with parameters or equations.
Futhermore,
they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.
We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
We also highlighted the substantial performance improvements our tools offer,
as this is crucial for construction and simulaiton of models of whole cells and organisms,
which necessitate the simulation of significantly large-scale models.

@@

1. The variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
2. It provides a namespace such that allows easily definition of multiple independent models in the same script,
3. It allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
4. It allows creatoion of instances which can be composed into a bigger model (@fig-composition).

@@

![Performance of different softwares to solve models from the cureted section of BioModels. (left) Run time for the model BIOMD3 as a function of the number of output points. (right) Run time for different models for 300 output points, using the geometric mean of the different softwares to order them.](src/performance/figures/performance.png){#fig-runtime}

>>>>

Chemical Reaction Networks (CRNs)
are a fundamental concept of modelling in numerous fields including systems biology, biochemistry, chemical engineering and epidemiology.
They are comprised of a set of chemical species or biological entities and representation of complex interaction and transformation between them through a series reactions.
These systems can be modelled and simulated through multiple approaches:
deterministic Ordinary Differential Equations (ODEs) to model macroscopic behavior,
Stochastic Differential Equations (SDEs) to model microscopic fluctuations,
and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
Instead of directly writing the equations for each of these formulations
which is error-prone and difficult to reuse,
these models can be written in a higher-level description
that can be compiled for these different types of simulations.

Several tools already exist for defining and simulating CRNs.
BioSimulators.org [@shaikhBioSimulatorsCentralRegistry2022],
a registry of simulation tools,
list at least 15 Python software including COPASI [@hoopsCOPASICOmplexPAthway2006], Tellurium [@choiTelluriumExtensiblePythonbased2018] and PySB [@lopezProgrammingBiologicalModels2013]. 
COPASI is a standalone software with a GUI for defining and simulating models.
It's widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow advanced scripting.
Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to simulate models.
PySB is a Python library
that created a DSL using standard Python to define models
which are then compiled to ODE using a Perl library called BioNetGen [@harrisBioNetGenAdvancesRulebased2016].

@@

Another challenge is the way models are written.
Many tools use a Domain Specific Language (DSL) for coding,
and support the System Biology Markup Language (SBML) as an exchange format [@huckaSBMLL3V2],
as direct SBML coding is impractical.
A DSL allows reuse of code in different programming environments,
but it will not be recognized by default in Integrated Development Environments (IDEs).
Therefore DSLs cannot provide the development experience such as syntax highlighting, code completion, refactoring, and static analysis,
unless such support is specifically developed.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
An extension for Visual Studio Code was developed, but its maintenance could be demanding task for the systems biology community.
PySB, using Python’s dynamic nature, created a DSL within Python.
By default, it uses global state to create species and parameters,
without assigning them to Python variables
or adding them explicitly to the model,
but this approach is not fully compatible with IDEs,
affecting the development experience.

To overcome these limitations,
we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows defining differential equation systems,
while SimBio builds on it for defining reaction networks.
They are focused on providing an ergonomic experience to end-users by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
Since they are coded in Python,
every part from model definition to simulation can be extended from Python syntax.
Being ?first-ever? pure Python packages for systems modelling,
they offer extensive extensibility,
from simple tasks like reusing integrators defined in other packages,
to complex ones like altering the compilation process to leverage some structure in the equations (For example...).
The models built using these packages can be introspected to create other representations,
such as graphs connecting species and/or reactions,
or tables with parameters or equations.
Furthermore,
they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.
We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
We also highlighted the substantial performance improvements our tools offer,
as this is crucial for construction and simulation of models of whole cells and organisms,
which necessitate the simulation of significantly large-scale models.

@@

1. The variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
2. It provides a namespace such that allows easily definition of multiple independent models in the same script,
3. It allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
4. It allows creation of instances which can be composed into a bigger model (@fig-composition).

@@

![Performance of different softwares to solve models from the curated section of BioModels. (left) Run time for the model BIOMD3 as a function of the number of output points. (right) Run time for different models for 300 output points, using the geometric mean of the different softwares to order them.](src/performance/figures/performance.png){#fig-runtime}
